---
title: "Lab 0b - Introduction to data"
output:
  html_document:
   html_document: default
#css: ../lab.css
highlight: pygments
theme: cerulean
pdf_document: default
---

The first step in data analysis is to summarize and describe the raw information - the data.  In this lab, you will gain insight into public opinion on energy issues and climate change by using a dataset from a survey given by the United Kingdom's Department of Energy and Climate Change to identify public attitudes towards energy and climate change. 

## Getting started

We begin by loading the data set of into the R workspace. After launching RStudio, load in the "Climate Change Public Tracker" data set that you've downloaded from Moodle and save it in a variable called "survey".

The data set `survey` that shows up in your workspace is a **data matrix**, with each row representing a **case** and each column representing a **variable**.  R calls this data format a **data frame**, which is a term that will be used throughout the labs.

To view the names of the variables, type the command

```{r names, eval=FALSE}
names(survey)
```

This returns quite a few variables, all of which are answer to survey questions. See the "survey question" document on Moodle to match up variable names with questions asked in the survey. 

**1.  Using information from tutorial 0a and the names function above, how many cases (measurements) are there in this data set?  How many variables?**

We can have a look at the first few entries (rows) of our data with the command

```{r head, eval=FALSE}
head(survey)
```

and similarly we can look at the last few by typing

```{r tail, eval=FALSE}
tail(survey)
```


You could also look at *all* of the data frame at once by typing its name into the console, but that might be unwise here.  We know `survey` has 2,121 rows, so viewing the entire data set would mean flooding your screen.  It's better to take small peeks at the data with `head`, `tail` or the subsetting techniques that you'll learn in a moment.

**2. Using the "head" or "tail" command, identify data types (e.g. categorical, discrete) represented in the dataset.**

*After determining data types, type in the command `str(survey)`. the `str` command displays the structure of the dataset, including the way that R classifies variables. R classifies variables as factors [synonymous with categories], characters [text objects], integers, numeric [includes continuous data], and logical [TRUE vs FALSE].*  

## Summaries and tables

 A good first step in any analysis is to distill all of that information into a few summary statistics and graphics.  As a simple example, the function `summary` returns a numerical summary: minimum, first quartile, median, mean, second quartile, and maximum. For `Number_Adults` this is

```{r, eval=FALSE}
summary(survey$Number_Adults)
```

**3. What can we glean from these summary numbers? Take a peek back at the variable description to give you context for what the variable is measuring.**

R also functions like a very fancy calculator.  If you wanted to compute the interquartile range for the Number Adults variable, you would look at the output from the summary command above and then enter

```{r weight-range-arith, eval=FALSE}
2 - 1
```

R also has built-in functions to compute summary statistics one by one.  For instance, to calculate the mean, median,  variance, and standard deviation of `Number_Adults`, type 

```{r weight-mean-var-median, eval=FALSE}
mean(survey$Number_Adults) 
median(survey$Number_Adults)
var(survey$Number_Adults)
sd(survey$Number_Adults)
```

While it makes sense to describe a quantitative variable like `Number_Adults` in terms of these statistics, what about categorical data?  We would instead consider the sample frequency or relative frequency distribution.  The function `table` does this for you by counting the number of times each kind of response was given. For example, to see the number of people who express concern about climate change:

```{r eval=FALSE}
table(survey$ClimChange_concerns)
```

or instead look at the relative frequency distribution by typing

```{r eval=FALSE}
table(survey$ClimChange_concerns)/2121 #Another way is to divide by length(survey$ClimChange_concerns)
```

Notice how R automatically divides all entries in the table by 2,121 in the command above. This is similar to something we observed in the last tutorial; when we multiplied or divided a vector with a number, R applied that action across entries in the vectors. As we see above, this also works for tables. Next, we make a bar plot of the entries in the table by putting the table inside the barplot command.

```{r, eval=FALSE}
barplot(table(survey$ClimChange_concerns), cex.names = 0.6, las = 2)
#The las = 2 function allows for all labels to be shown on the x-axis
```

Notice what we've done here! We've computed the table of `survey$ClimChange_concerns` and then immediately applied the graphical function, `barplot`. This is an important idea: R commands can be nested. You could also break this into two steps by typing the following:

```{r eval=FALSE}
CC_concern <- table(survey$ClimChange_concerns)

barplot(CC_concern, las = 2, cex.names = 0.6)
```

Here, we've made a new object, a table, called `CC_concern` (the contents of which we can see by typing `CC_concern` into the console) and then used it in as the input for `barplot`. The special symbol `<-` performs an *assignment*, taking the output of one line of code and saving it into an object in your workspace (you can use = or <- interchangably in performing an assignment).   This is another important idea that we'll return to later.

**4.  Create a numerical summary for `Number_people` and compute the interquartile range. Compute the relative frequency distribution for `Area` and `Working_status`. How many rural respondants are there? What proportion of the sample reports being working full time?**

The `table` command can be used to tabulate any number of variables that you provide.  For example, to examine participants' level of concern about climate change across each age group, we could use the following.

```{r eval=FALSE}
table(survey$Age,survey$ClimChange_concerns)
```

To create a mosaic plot of this table, we would enter the following command.

```{r eval=FALSE}
mosaicplot(table(survey$Age,survey$ClimChange_concerns), las = 1)
```

We could have accomplished this in two steps by saving the table in one line and applying `mosaicplot` in the next (see the table/barplot example above).

**5.  What does the mosaic plot reveal about age and climate change concern?**

## Interlude: How R thinks about data

We mentioned that R stores data in data frames, which you might think of as a type of spreadsheet. Each row is a different observation (a different respondent) and each column is a different variable (the first is `respnum`, the second `Energy_saving_thought` and so on). We can see the size of the data frame next to the object name in the workspace or we can type

```{r dim, eval=FALSE}
dim(survey)
```

which will return the number of rows and columns. Now, if we want to access a subset of the full data frame, we can use row-and-column notation. For example, to see the sixth variable of the 567<sup>th</sup> respondent, use the format

```{r cdc-row567-column6, eval=FALSE}
survey[567,10]
```

which means we want the element of our data set that is in the 567<sup>th</sup> row (meaning the 567<sup>th</sup> person or observation) and the 10<sup>th</sup> column (in this case, weight). You will notice in the output that "Levels" are provided. These are the various levels of the `ClimateChange_Challenge` variable, or the different potential values of this variable.  We know that `ClimateChange_Challenge` is the 6<sup>th</sup> variable because it is the 10<sup>th</sup> entry in the list of variable names

```{r names-again, eval=FALSE}
names(survey)
```


To see the whether the first 10 respondents view climate change as a challenge, we can type

```{r, eval=FALSE}
survey[1:10,10]
```

In this expression, we have asked just for rows in the range 1 through 10.  R uses the `:` to create a range of values, so 1:10 expands to 1, 2, 3, 4, 5, 6, 7, 8, 9, 10. You can see this by entering

```{r numbers-1to10, eval=FALSE}
1:10
```

Finally, if we want all of the data for the first 10 respondents, type

```{r first-10-rows, eval=FALSE}
survey[1:10,]
```

By leaving out an index or a range (we didn't type anything between the comma and the square bracket), we get all the columns. When starting out in R, this is a bit counterintuitive. As a rule, we omit the column number to see all columns in a data frame. Similarly, if we leave out an index or range for the rows, we would access all the observations, not just the 567<sup>th</sup>, or rows 1 through 10. Try the following to see the weights for all 2,121 respondents fly by on your screen

```{r 6th-column, eval=FALSE}
survey[,10]
```

An alternative method to access the Climate Change Challenge is by referring to the name.  Previously, we typed `names(survey)` to see all the variables contained in the  data set. We can use any of the variable names to select items in our data set. 

```{r weight, eval=FALSE}
survey$ClimateChange_Challenge
```

The dollar-sign tells R to look in data frame `survey` for the column called `ClimateChange_Challenge`.  Since that's a single vector, we can subset it with just a single index inside square brackets.  We see the Fracking support response for the 567<sup>th</sup> respondent by typing

```{r, eval=FALSE}
survey$ClimateChange_Challenge[567]
```

Similarly, for just the first 10 respondents

```{r, eval=FALSE}
survey$ClimateChange_Challenge[1:10]
```

The command above returns the same result as the `survey$ClimateChange_Challenge[1:10,6]` command. Both row-and-column notation and dollar-sign notation are widely used, which one you choose to use depends on your personal preference.

## A little more on subsetting

It's often useful to extract all individuals (cases) in a data set that have specific characteristics. We accomplish this through *conditioning* commands.  First, consider expressions like

```{r, eval=FALSE}
survey$Area == "Rural"
```

or

```{r true-over30, eval=FALSE}
survey$Working_status == "Full time"
```

These commands produce a series of `TRUE` and `FALSE`  values. There is one value for each respondent, where `TRUE` indicates that the person is from a rural area (first command) or works full time (second command).

Suppose we want to extract just the data for the rural areas, or just for those who work full time. We can use the R function `subset` to do that for us. For example, the command

```{r males, eval=FALSE}
rural <- subset(survey, survey$Area == "Rural")
```

will create a new data set called `rural` that contains only the rural responses from the `survey` data set. In addition to finding it in your workspace alongside its dimensions, you can take a peek at the first several rows as usual

```{r, eval=FALSE}
head(rural)
```

This new data set contains all the same variables but just about 25% of the rows. It is also possible to tell R to keep only specific variables, which is a topic we'll discuss in a future lab. For now, the important thing is that we can carve up the data based on values of one or more variables.

As an aside, you can use several of these conditions together with `&` and `|`. The `&` is read "and" so that

```{r, eval=FALSE}
rural_and_over2Adults <- subset(survey, survey$Area == "Rural" & survey$Number_Adults > 2)
```

will give you the data for respondents from rural regions with more than 2 adults in the household. The `|` character is read "or" so that

```{r, eval=FALSE}
rural_or_over2Adults <- subset(survey, survey$Area == "Rural" | survey$Number_Adults > 2)
```

will take respondents from rural regions or those with greater than two adults in the household (why that's an interesting group is hard to say, but right now the mechanics of this are the important thing). In principle, you may use as many "and" and "or" clauses as you like when forming a subset.

**6.  Create a new object called `Income16000_and_renewables` that contains all observations of respondents making greater than 16000 pounds that strongly support solar energy.**

## Quantitative data

With our subsetting tools in hand, we'll now return to the task of the day: making basic summaries of the survey.  We've already looked at categorical data so now let's turn our attention back to quantitative data.  Two common ways to visualize discrete quantitative data are with box plots and histograms.  We can construct a box plot for a single variable with the following command.

```{r boxplot-height, eval=FALSE}
boxplot(survey$Income)
```

You can compare the locations of the components of the box by examining the summary statistics.

```{r summary-height, eval=FALSE}
summary(survey$Income)
```

Confirm that the median and upper and lower quartiles reported in the numerical summary match those in the graph. The purpose of a boxplot is to provide a thumbnail sketch of a variable for the purpose of comparing across several categories. So we can, for example, compare the response to "Is climate change among top three challenges facing Britain today?" across income levels.  

```{r, eval=FALSE}
boxplot(survey$Income ~ survey$ClimateChange_Challenge, cex.axis = 0.4)
```

The notation here is new. The `~` character can be read *versus* or *as a function of*. So we're asking R to give us a box plots of number of people in the household where the groups are defined by their fracking support.  

**7.  What does this box plot show? Pick another categorical variable from the data set and see how it relates to Income.** 

Finally, let's make some histograms. We can look at the histogram for the income of our respondents with the command

```{r hist-age, eval=FALSE}
hist(survey$Income)
```

Histograms are generally a very good way to see the shape of a single distribution, but that shape can change depending on how the data is split between the different bins.  You can control the number of bins by adding an argument to the command. In the next two lines, we first make a default histogram of `Income` and then one with 3 breaks.

```{r hist-bmi, eval=FALSE}
hist(survey$Income)
hist(survey$Income, breaks = 3)
```

Note that you can flip between plots that you've created by clicking the forward and backward arrows in the lower right region of RStudio, just above the plots.  How do these two histograms compare?

At this point, we've done a good first pass at analyzing the information in the survey. We've also picked up essential computing tools -- summary statistics, subsetting, and plots -- that will serve us well throughout this course.

* * *

## On Your Own

-   Come up with a research question of your choosing and explore it with the data provided.  
+  Produce at least one figure and evaluate your question (make sure to annotate own code)

Once you are done with the "On your own", knit a markdown document (using the "Knit" icon at the top of the R markdown console). Save a version of your R markdown document in your Google Drive, so that you can use it as a future resource. 

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). This lab was adapted for BIOL 374 by Chantal Koechli from a lab adapted for OpenIntro by Andrew Bray and Mine &Ccedil;etinkaya-Rundel from a lab written by Mark Hansen of UCLA Statistics.
</div>